---
title: "AI_Analysis_NoCode"
author: "Aarul"
date: "2025-12-06"
output: html_document
---

# Final Project Report: AI's impact on the Job Market
## Collaborators: Aarul Motiani, Eshaan Garg, Wayne Zhao, Finn Caughie
================================================================================================================================================================
## AARUL'S RESEARCH QUESTION: How has demand for AI-related roles grown over the past 5 years?

'''{r, echo=FALSE}
## Loading libraries and datasets
```{r}
library(tidyverse)
library(ggplot2)
library(lubridate)

## Aarul's computer loaded datasets
job_postings <- read_delim("/Users/Aarul/UW/INFO 201/Final Project/INFO201_Final_Project/AI_Job_Posting_Trends.csv")
salary_data <- read_delim("/Users/Aarul/UW/INFO 201/Final Project/INFO201_Final_Project/data_science_salaries.csv")
```
## Data exploration for Aarul's datasets
```{r}
head(job_postings)
names(job_postings)
nrow(job_postings)
ncol(job_postings)


head(salary_data)
names(salary_data)
nrow(salary_data)
ncol(salary_data)
```

## Dataset 1 (job_postings) cleaning and organizing
```{r}
job_postings_clean <- job_postings %>%
  mutate(posting_date = as.Date(posting_date, format = "%m/%d/%y")) %>%
  filter(!is.na(posting_date)) %>%
  distinct()
head(job_postings_clean)
```

## Dataset 2 (market_insights) cleaning and filtering data
```{r}
salary_data_clean <- salary_data %>%
  filter(work_year >= 2020) %>%
  filter(work_year <= 2023) %>%
  filter(!is.na(salary_in_usd)) %>%
  distinct()
head(salary_data_clean)

```

## Grouping and summary statistics for job postings per year
```{r}
jobs_per_year <- job_postings_clean %>%
  mutate(year = format(posting_date, "%Y")) %>%
  group_by(year) %>%
  summarize(job_count = n())
head(jobs_per_year)
```

## Grouping and adding columnn for average salary per year calculation
```{r}
salary_per_year <- salary_data_clean %>%
  rename(year = work_year) %>%
  mutate(year = as.character(year)) %>%
  group_by(year) %>%
  summarize(avg_salary = mean(salary_in_usd), median_salary = median(salary_in_usd))

head(salary_per_year)
```

## Creating numerical variables for data analysis (using lag function)
```{r}
salary_per_year <- salary_per_year %>%
  mutate(salary_category = "AI/Data Science") %>%
  arrange(year) %>%
  mutate(prev_salary = lag(avg_salary), salary_growth_rate = (avg_salary - prev_salary) / prev_salary * 100)

print(salary_per_year)
```


## Merging both the datasets
```{r}
merged <- jobs_per_year %>%
  left_join(salary_per_year, by = "year")

head(merged)
```


## Research question data analysis to calculate salary growth
```{r}
first_year <- salary_per_year %>% filter(year == "2020")
last_year <- salary_per_year %>% filter(year == "2023")
total_salary_growth <- ((last_year$avg_salary - first_year$avg_salary) / first_year$avg_salary) * 100
```

## Company size grouping and evaluation
```{r}
company_summary <- job_postings_clean %>%
  group_by(company_size) %>%
  summarize(total_jobs = n())

print(company_summary)
```

## Visualizations
```{r}
salary_per_year %>% 
ggplot(aes(x = as.numeric(year), y = avg_salary)) +
  geom_line(size = 1.5, color = "black") +
  geom_point(size = 5, color = "red") +
  labs(title = "Average AI Job Salary Growth (2020-2023)", x = "Year", y = "Average Salary (USD)")
```

```{r}
ggplot(company_summary, aes(x = company_size, y = total_jobs, group = 1)) +
  geom_line(size = 1.5, color = "black") +
  geom_point(size = 5, color = "red") +
  labs(title = "AI Job Postings by Company Size (2020-2023)", x = "Company Size", y = "Number of Job Postings") 
```

================================================================================================================================================================

## FINN'S RESEARCH QUESTION (2): Which industries have experienced the greatest increase in AI-related job openings?

# Loading and checking data
```{r}
insights <- read_delim("/Users/Aarul/UW/INFO 201/Final Project/INFO201_Final_Project/ai_job_market_insights.csv")
head(insights)
```


```{r}
trends <- read_delim("/Users/Aarul/UW/INFO 201/Final Project/INFO201_Final_Project/AI_Job_Posting_Trends 2.csv")
head(trends)
```

#Cleaning the data 
```{r}
insights_clean <- insights %>%
  distinct()

trends_clean <- trends %>%
  mutate(posting_date = as.Date(posting_date, format = "%m/%d/%y")) %>%
  distinct()
```

#Updating data so I can use it for merging and graphing
```{r}
trends_per_month <- trends_clean %>%
  mutate(year = format(posting_date, "%Y"), month = format(posting_date, "%m")) %>%
  group_by(industry, year, month) %>%
  summarize(postings_count = n()) #counts number of rows per group
```

```{r}
industry_increase <- trends_per_month %>%
  group_by(industry) %>%
  summarize(first = first(postings_count), last = last(postings_count), increase = last - first) %>%
  arrange(desc(increase))
```

```{r}
head(trends_per_month)
head(industry_increase)
```

#Merging the data
```{r}
merged <- merge(industry_increase, trends_per_month, by = "industry")
head(merged)
```

```{r}
merged <- merged %>%
  arrange(desc(increase))

```

#This graph shows the job postings for the top 5 growing industries over the years
```{r}
yearly_increase <- merged %>%
  group_by(industry, year) %>%
  summarize(yearly_total = sum(postings_count)) %>%
  arrange(industry, year)

top5 <- merged %>%
  distinct(industry, increase) %>%
  arrange(desc(increase)) %>%
  slice(1:5) %>%
  pull(industry)

yearly_increase_top5 <- yearly_increase %>%
  filter(industry %in% top5)

ggplot(yearly_increase_top5, aes(x = as.numeric(year), y = yearly_total, color = industry)) +
  geom_line(linewidth = 1.1) +
  labs(
    title = "AI Job Postings for Top 5 Growing Industries Per Year",
    x = "Year",
    y = "Number of Postings"
  )
```


#This graph uses merged data to show increase of AI job postings by industry per year 
```{r}
ggplot(merged, aes(x = industry, y = increase, fill = industry)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Increase in AI Job Postings by Industry",
    x = "Industry",
    y = "Increase in Job Postings"
  )
```

================================================================================================================================================================
## WAYNE'S RESEARCH QUESTION: How has AI-related hiring shifted in non-tech sectors?

## Cleaning and Merging
```{r}
lowercase_clean <- function(df) {
  names(df) <- tolower(names(df))
  df
}

job_postings <- job_postings %>%
  rename(location = company_location)

insights <- lowercase_clean(insights)

names(job_postings)
names(insights)

merged_data <- job_postings %>%
  left_join(insights, by = "job_title", relationship = "many-to-many")

merged_data <- merged_data %>%
  select(-ends_with(".y"))

merged_data <- merged_data %>%
  mutate(
    remote_friendly = ifelse(remote_category == "Fully Remote", "Yes", remote_friendly),
    remote_friendly = ifelse(remote_category == "On-site", "No", remote_friendly),
    remote_friendly = ifelse(remote_category == "Hybrid", "Hybrid", remote_friendly)
      )

merged_data
```


## Sort data by the amount of times a job got labeled "Growth" or "Decline"
```{r}
growth_count <- insights %>%
  filter(!is.na(job_growth_projection)) %>%
  count(job_title, job_growth_projection) %>%
  group_by(job_title) %>%
  mutate(total = sum(n))
```

## Top jobs with the most amount of "Growth" labels
```{r}
top_jobs <- growth_count %>%
  filter(job_growth_projection == "Growth") %>%
  arrange(desc(n)) %>%
  head(10)
```

## Industries with the most job postings

```{r}
insights %>%
  count(industry) %>%
  ggplot(aes(x = reorder(industry, n), y = n)) +
  geom_col(fill = "coral") +
  coord_flip() +
  labs(
    title = "Job Postings by Industry",
    x = "Industry",
    y = "Number of Postings"
  )
```

## Visualization of AI Job Roles with the most "Growth" labels by percentage

```{r}
growth_percent <- insights %>%
  filter(!is.na(job_growth_projection)) %>%
  count(job_title, job_growth_projection) %>%
  group_by(job_title) %>%
    mutate(total = sum(n)) %>%
    mutate(percentage = (n / total) * 100) %>%
  filter(job_growth_projection == "Growth", total >= 2) %>%
  arrange(desc(percentage)) %>%
  head(10)

ggplot(growth_percent, aes(x = reorder(job_title, percentage), y = percentage)) +
  geom_col(fill = "#3498db") +
  coord_flip() +
  ylim(0, 60) +
  labs(
    title = "Top 10 AI Job Roles with Highest Growth Rate",
    x = "Job Role",
    y = "% of Job Postings Labeled Growth"
  )

```
## Visualization of Job Growth Trajectory VS AI Adoption Level
```{r}
growth_by_adoption <- insights %>%
  filter(!is.na(job_growth_projection), !is.na(ai_adoption_level)) %>%
    count(job_title, ai_adoption_level, job_growth_projection) %>%
    filter(job_growth_projection == "Growth") %>%
    arrange(ai_adoption_level, desc(n)) %>%
    group_by(ai_adoption_level) %>%
    slice_max(n, n = 5) %>%
    ungroup()
 
  ggplot(growth_by_adoption, aes(x = ai_adoption_level, y = n, color = job_title, group = job_title)) +
    geom_line() +
    geom_point(size = 2) +
    scale_x_discrete(limits = c("Low", "Medium", "High")) +
    labs(
      title = "Job Growth Trajectory by AI Adoption Level",
      x = "AI Adoption Level",
      y = "Number of Growth Postings",
      color = "Job Role"
    )
```
================================================================================================================================================================
## ESHAAN'S RESEARCH QUESTION: What countries are leading in AI job growth relative to GDP or population?

## Load libraries
```{r}
library(tidyverse)
library(readr)
library(dplyr)
library(ggplot2)
```

## Load datasets for Question 4
```{r}
aiJobs <- read_csv("/Users/Aarul/UW/INFO 201/Final Project/INFO201_Final_Project/data_science_salaries.csv")
countries <- read_csv("/Users/Aarul/UW/INFO 201/Final Project/INFO201_Final_Project/countries of the world - Eshaan.csv")

head(aiJobs)
head(countries)
```

## Clean countries dataset
```{r}
countriesClean <- countries %>%
  mutate(
    Country = str_trim(Country),
    Population = as.numeric(gsub("[^0-9]", "", Population)),
    GDP_per_capita = as.numeric(gsub("[^0-9]", "", `GDP ($ per capita)`)),
    Total_GDP = GDP_per_capita * Population
  ) %>%
  select(Country, Region, Population, GDP_per_capita, Total_GDP) %>%
  filter(!is.na(Population), !is.na(Total_GDP))

```

## Count AI jobs by country
```{r}
aiByCountry <- aiJobs %>%
  count(employee_residence, name = "jobCount") %>%
  arrange(desc(jobCount))

head(aiByCountry, 20)
```

## Check which countries won't match
```{r}
aiByCountry %>%
  filter(!employee_residence %in% countriesClean$Country) %>%
  select(employee_residence, jobCount)
```

## Merge datasets and calculate metrics
```{r}
finalData <- aiByCountry %>%
  inner_join(countriesClean, by = c("employee_residence" = "Country")) %>%
  mutate(
    jobsPerMillion = (jobCount / Population) * 1e6,
    jobsPerBillionGdp = (jobCount / (Total_GDP / 1e9))
  ) %>%
  rename(Country = employee_residence)

nrow(finalData)
head(finalData)
```

## Visualization 1: Top countries by absolute job count
```{r}
finalData %>%
  slice_max(jobCount, n = 15) %>%
  ggplot(aes(x = reorder(Country, jobCount), y = jobCount)) +
  geom_col(fill = "#2E86AB") +
  coord_flip() +
  labs(
    title = "Top 15 Countries by AI Job Postings (2019-2024)",
    x = NULL,
    y = "Number of Jobs"
  ) +
  theme_minimal()
```

## Visualization 2: Jobs per million population
```{r}
finalData %>%
  slice_max(jobsPerMillion, n = 15) %>%
  ggplot(aes(x = reorder(Country, jobsPerMillion), y = jobsPerMillion)) +
  geom_col(fill = "#A23B72") +
  coord_flip() +
  labs(
    title = "Top 15 Countries by AI Jobs per Million Population",
    x = NULL,
    y = "Jobs per Million People"
  ) +
  theme_minimal()
```

## Visualization 3: Jobs per billion GDP
```{r}
finalData %>%
  filter(jobsPerBillionGdp < Inf) %>%
  slice_max(jobsPerBillionGdp, n = 15) %>%
  ggplot(aes(x = reorder(Country, jobsPerBillionGdp), y = jobsPerBillionGdp)) +
  geom_col(fill = "#F18F01") +
  coord_flip() +
  labs(
    title = "Top 15 Countries by AI Jobs per Billion USD GDP",
    x = NULL,
    y = "Jobs per Billion USD"
  ) +
  theme_minimal()
```

## Summary statistics table
```{r}
finalData %>%
  arrange(desc(jobCount)) %>%
  select(Country, jobCount, jobsPerMillion, jobsPerBillionGdp) %>%
  mutate(
    jobsPerMillion = round(jobsPerMillion, 1),
    jobsPerBillionGdp = round(jobsPerBillionGdp, 2)
  ) %>%
  head(15)
```

================================================================================================================================================================

# 1. Abstract
----------------------------------------------------------------------------------------------------------------------------------------------------------------
### In this paper we analyze the growth in AI-related positions from 2020 through 2023, using both job postings and salary data from Kaggle. The purpose of our paper is to answer four research questions; (1) How has demand for AI-related roles grown over the past 5 years? (2) Which industries have experienced the greatest increase in AI-related job openings? (3) How has AI-related hiring shifted in non-tech sectors? (4) What countries are leading in AI job growth relative to GDP or population? We find that there was a 47.5% increase in the average salary of people working in AI jobs over this period, from an average of $102,251 in 2020 to $150,791 in 2023. This indicates that employers continue to be highly interested in acquiring AI talent. Our results also indicate a fairly even distribution of hiring activity among all firm size categories as 16,742 AI related job postings were placed by larger firms versus 16,571 by smaller firms. In addition, our analysis of hiring by sector indicates that each sector has had different rates of adopting AI technology and we also identify the presence of concentrations in markets for AI talent.

# 2. Introduction
----------------------------------------------------------------------------------------------------------------------------------------------------------------
## Background Context:

### AI has evolved dramatically from a laboratory experiment to a major source of business innovation and growth. In the last five years, the rate of AI adoption has grown significantly across many industries, and the rate of change is transforming workforce skills and needs. The increasing investment in AI capabilities at the company level has created unprecedented demand for individuals with the ability to develop and deploy AI systems. The change is occurring across virtually all sectors including, healthcare, financial services, manufacturing and retail. All of these sectors are now using AI within their operations.

### The AI job market represents a key point of convergence between technological innovation and economic opportunity. Organizations competing for the most capable professionals to develop, build, and use machine learning, natural language processing, and computer vision systems are increasingly aggressive in their pursuit of such talent. This competition is evident through increases in salary levels, greater numbers of job postings, new specialization opportunities and changing expectations regarding required skills. Research on previous studies documenting the broad increase in number of technology jobs and the establishment of data science as a recognized profession, however, has not examined in detail the specifics of AI related professional roles, including trends in compensation, geographic distribution, industry adoption, and organizational preference.

### Understanding these changes has significant implications for several types of stakeholders. Job seekers will benefit from having access to data that informs career decisions and investments in skill development. Educational institutions will be able to create curricula aligned with actual market demand. Policymakers will have access to information needed to inform workforce development initiatives. Finally, business leaders will have benchmarking tools available to support their strategic efforts to acquire talented employees. Our analysis will provide this type of information by analyzing a comprehensive set of data spanning the period from 2020 to 2023; a time during which AI was advancing at a very high rate, and was being adopted by organizations at an increasingly rapid pace.

## Research Objectives: 

## Research Question 1 (Aarul Motiani): How has demand for AI-related roles grown over the past 5 years?
### Demand can be measured using two main ways; job postings and salaries. Since we have analyzed 50,000 job postings and salary data, we can find out if an increase in employers' ability to pay employees more money is evidence that there is actually more demand in the labor market for AI skills. Both of these elements address the amount of and the quality of the demand.

## Research Question 2 (Finn Caughie): Which industries have experienced the greatest increase in AI-related job openings?
### Job postings data from several months will help us to see the trend of which industries are developing AI related hiring practices. We will also determine if AI growth is limited to tech companies or is more widely spread throughout various other industries. This research will also help us to understand the type of industries that are making the heaviest investments into their AI capabilities.

## Research Question 3 (Wayne Zhao): How has AI-related hiring shifted in non-tech sectors?
### The purpose of this section is to look at industries that were traditionally outside the scope of the technology sector and better to understand the level of AI adoption within those industries. This section will address whether AI-related positions are still specialized or have been adopted as common across all areas of the business. The sections on adoption rates, the likelihood of job automation and work-at-home arrangements will be used to study how non-tech industries have developed unique approaches to acquiring AI talent.

## Research Question 4 (Eshaan Garg): What countries are leading in AI job growth relative to GDP or population?
### This section will be used to compare the number of AI-related jobs that exist in each country in relation to its population size or the size of the economy. This comparison will allow us to differentiate between large economies that may have an enormous number of total jobs available, versus small countries that have a significantly larger proportion of the workforce with AI related skills. These comparisons will help us to understand the global distribution of AI expertise.

## Why this work matters: 

### Job Seekers/Career Planners: Understanding where wages are increasing, where there is an abundance of job openings, and what areas have growing populations, allows Career Planning and Skill Development to be done strategically. The documented 47.5% salary increase shows how much money one can make when investing in AI knowledge; Additionally, industry specific findings will allow individuals to find which industry has the most available jobs.

### Educational Institutions: Educational Institutions (Universities/Training Programs) will be able to design their curriculum so it matches the job requirements of companies in the marketplace. It also will give Educators a way to tell their students about the job opportunities that exist today, as opposed to what gets published in the press. The even distribution of job opportunities across all company sizes implies that graduates will have the skills to enter into many different types of organizational settings.

### Policymakers/Economic Development Officials: Policymaker's/Economic Development Official's ability to understand how the talent pool of AI is geographically distributed, will enable them to create workforce development programs and develop economic policies that promote the creation of AI talent. When analyzing the top performing countries in AI Job Growth as compared to their Gross Domestic Product, Policymakers will be able to see successful models of AI Talent Development.

### Business Leaders/HR Professionals: Business leaders who are recruiting AI talent will require information based upon data regarding their compensation package, competitors' hiring practices, and availability of qualified candidates. Data related to the distribution of job openings by Company Size and Industry Trends will assist businesses to position themselves in terms of their recruiting strategy. In addition, the data regarding salary growth will provide businesses with concrete data to compare against other companies in their respective industry.

### Researchers/Economists: The systematic nature of our research will contribute to the academic study of the impact of technology on the labor markets. We provide empirical evidence for the theories regarding skill-biased technological change and the future of work.

### Overall, the significance of this Report is that it provides a comprehensive, data driven perspective of a rapidly changing Labor Market Segment. Unlike anecdotal reports, we analyzed over 100,000 job postings and salary records to determine consistent patterns. The multi-faceted analysis of the job market (temporal, industrial, organizational, and geographical) provides a level of detail that goes beyond the typical headline statistic. As AI represents a fundamental change in how businesses operate and compete, the AI Talent Market is a leading indicator of overall Economic Change.

# 3. Data Sources
----------------------------------------------------------------------------------------------------------------------------------------------------------------

## AI Jobs Posting Trends from 2019–2023 Source: 
## Kaggle dataset: AI Jobs Posting Trends from 2019–2023. https://www.kaggle.com/datasets/pranjali1121/ai-jobs-posting-trends-from-2019-2023 

## Who collected the data: This dataset was uploaded by the user pranjali1121 and is on Kaggle which is a platform where community members can post datasets for people to use freely. Usage permissions: The dataset is publicly available on Kaggle. It can be freely accessed and downloaded by anyone with a Kaggle account. There are no restricted usage terms outside of the standard Kaggle terms (not for redistribution outside Kaggle without proper citation). 

## What it represents: It shows the trends in AI-related job postings from 2019–2023, showing how demand for AI skills has changed over time. The data typically includes counts or indicators of job postings by year. 

## Why we chose it: This dataset directly supports research questions about how AI job demand has changed over time, giving us a view of hiring trends and helping analyze growth patterns in the AI labor market. 

## AI-Powered Job Market Insights Source: 

## Kaggle dataset: AI-Powered Job Market Insights. https://www.kaggle.com/datasets/uom190346a/ai-powered-job-market-insights 

## Who collected the data: This dataset was provided by the user uom190346a. 

## Usage permissions: It’s freely accessible on Kaggle for anyone to download and use based on Kaggle’s standard terms. There are no restrictions (non-commercial only) listed on the page. 

## What it represents: The dataset has detailed job listings with things like job title, industry, company size, AI adoption level, automation risk, required skills, salary ranges, and growth projection which was useful for understanding AI-related jobs. 

## Why chosen: This dataset supports our research questions about current AI job market characteristics, such as what skills are in demand, what industries are hired, and how job attributes vary with AI adoption. 

## Countries of the World Dataset Source: 

## Kaggle dataset: Countries of the World https://www.kaggle.com/datasets/fernandol/countries-of-the-world 

## Who collected the data: The dataset was uploaded by the Kaggle user fernandol, who compiled country-level statistics originally sourced from the CIA World Factbook. 

## Usage permissions: The dataset is publicly available and free to download from Kaggle. Anyone with a Kaggle account can access it, with no special restrictions beyond standard Kaggle usage and citation rules. 

## What it represents: This dataset includes demographic, economic, and geographic indicators for countries around the world such as population, area, GDP, literacy rates, and other statistics. 

## Why chosen: This dataset helps support research questions related to comparing AI job growth across countries, especially when adjusting for factors like GDP or population size. It provides the baseline national statistics needed to standardize and compare AI job expansion globally.

----------------------------------------------------------------------------------------------------------------------------------------------------------------
# 4. Data

## Insights data: 500 rows and 10 columns Each row represents one individual job posting, describing its role, industry, location, required skills, salary, AI adoption level, and other characteristics. In other words, every row is a single job listing with all its attributes recorded. 

## The relevant variables are those that help track AI job growth over time, across industries, and countries. Variables like posting_date, year, job_title, and job_category are coded as dates or categorical text to identify AI roles and measure demand. Industry and location are categorical and allow comparisons across sectors and countries, while ai_adoption_level, automation_risk, required_skills, and job_growth_projection are categorical variables that show patterns in non-tech sectors. Salary_usd is numerical and provides economic context for the trends. 

## There are no missing values in any of the columns, as all counts are zero. There are also no negative or otherwise implausible salaries, and the salary values fall within a reasonable range (approximately $32,000 to $155,000). The categorical variables (AI_Adoption_Level, Automation_Risk, Job_Growth_Projection, Company_Size) only contain expected, consistent categories, and there are no duplicate rows in the dataset.

----------------------------------------------------------------------------------------------------------------------------------------------------------------

# 5. Method & Results 

## Data Cleaning and Processing Choices 

### To make sure the datasets had the appropriate number of rows and columns for analysis, we explored all datasets. The datasets needed cleaning and reorganization, so we took many steps in preparing the data to be processed. These included removing duplicate entries with distinct(), converting date fields into real date formats using as.Date(), and extracting year and month values with format(). We standardize column names on market_insights.csv and job_postings.csv by lowercasing them and renaming inconsistent variables to ensure compatibility across datasets. For the AI job trends data, we also simplified categorical variables such as company_size (S, M, L) and remote_category (Yes, No, Hybrid) to improve interpretability. Additionally, for the countries dataset, we removed commas using gsub() and converted population and GDP values to numeric using as.numeric(). We also used functions like filter(), group_by(), mutate(), and summarize() to organize job posting data by industry, job title, and time period. We counted total monthly job postings, calculated increases by month, and created summary measures such as total job count, job per million population, and jobs per billion USD GDP. We merged multiple datasets through left_join(), using keys such as industry, job_title, and employee_residence to align job postings data with market insights and country-level demographic data. Left_join() was used to minimize data loss because a dataset had more complete observations without needing much cleaning while the other had less. When we merged the countries dataset, we matched 15 out of 20 countries, excluding 5 due to missing data or mismatching names. 

## Technical Decisions

## There were many times when we had to make technical decisions for missing data. We decided to remove all rows containing NA in necessary columns like Population or Total_GDP because they are important for computation. Mismatched country names that prevented joining were also removed from the analysis. Missing dates or undefined job titles were filtered out of the job posting dataset. However, missing salary values were not replaced as it could have caused later issues if we used it to compute growth.

## Variables Relevant to Research Questions 

### We used many different variables to answer the research questions. For studying demand and salary growth, we relied on company_size, salary_usd, year, and industry. For growth projections, industry comparisons and the growth of AI job roles, we used job_title, ai_adoption_level, and number of postings. Country-level data, employee_residence, Population, and Total_GDP and the countries dataset itself were essential for finding trends across regions and seeing which countries led in AI job growth. 

## Steps Taken to Conduct Analysis 

### Our first step was confirming the number of rows, columns, and how well the variables/column names worked together to answer the research questions. Then we cleaned each dataset by removing duplicates, standardizing the column names, and cleaning dates using as.Date() to make it usable and format() to extract months and years. Next, we grouped job postings by month and counted them using summarize(n()). We calculated salary growth using lag() and mutate(), allowing for comparison across years. To merge information across the different datasets, we used left_join() most often to combine posting trends with market insights, and demographic datasets. We identified the top 10 AI job roles by selecting titles with the highest growth projections and computing their share of total postings. For industry trends, we merged datasets based on industry names and filtered the top five industries to visualize their trajectories. We then created visualizations including bar, line, and comparative plots to show the trends in the growth of industries, salary changes, AI adoption levels, and overall, AI job patterns. Finally, after merging the jobs dataset with the “countries” dataset, I computed normalized metrics to analyze international hiring patterns. 

## Challenges and Limitations 

### Our main challenge was data inconsistencies and reliability. Five countries could not be matched during merging due to inconsistent naming conventions or missing demographic data, reducing the completeness of the international comparison. Some job postings lacked reliable salary information, which limited the accuracy of salary growth calculations. Additionally, low or uneven AI adoption reporting across industries created gaps in comparing demand versus readiness. Sector categories were broad and did not fully capture sub-industry differences. Finally, merging datasets from different sources required resolving inconsistencies in naming, capitalization, and granularity, which caused minor errors. 

## Research Q1: How has demand for AI-related roles grown over the past 5 years? 

## Data / Visualization: 

```{r}
salary_per_year %>% 
ggplot(aes(x = as.numeric(year), y = avg_salary)) +
  geom_line(size = 1.5, color = "black") +
  geom_point(size = 5, color = "red") +
  labs(title = "Average AI Job Salary Growth (2020-2023)", x = "Year", y = "Average Salary (USD)")
```

```{r}
ggplot(company_summary, aes(x = company_size, y = total_jobs, group = 1)) +
  geom_line(size = 1.5, color = "black") +
  geom_point(size = 5, color = "red") +
  labs(title = "AI Job Postings by Company Size (2020-2023)", x = "Company Size", y = "Number of Job Postings") 
```

### To measure the overall demand, we merged yearly job posting counts with annual salary averages from 2020-2023. We split the results using 2 visualizations, one involving year-over-year growth in average AI salaries, and another showing the distribution of job postings across company sizes. These in combination help us analyze the quality and intensity of the markets' demand for AI-related roles. 

## Findings: 

### The line graph of average salary growth demonstrates a rise in salary for AI (47.5% increase) and its related roles, showing market demand and employers’ increasing willingness to pay employees with AI expertise. Furthermore, our company vs size comparison shows that demand is distributed across organizations regardless of size. Larger companies posted 16,742 AI-related roles while smaller companies posted 16,571, which is a minimal difference. Suggesting that AI adoption and hiring are not only exclusive to big tech companies, but small companies are also adamant in recruiting AI workers and involving themselves in the world of AI. 

## Limitations:  

### Despite clear patterns, the analysis may understate real hiring levels. Job posting datasets reflect only publicly advertised positions and exclude internal hiring channels, contract platforms, or informal recruitment. Salary data may also be skewed toward countries and industries that post compensation more transparently. In addition, the dataset covers only 2020–2023, limiting our ability to determine whether recent trends represent short-term acceleration or long-term structural change. 

## Potential Improvements: 

### Future research could incorporate additional job board datasets, extend the time range for trend analysis, and include skill-specific breakdowns (e.g., demand for machine-learning engineers vs. data analysts). Adding measures of seniority and job type would also help distinguish between growth in entry-level roles versus high-skill specialist positions. 

## Research Q2: Which industries have experienced the greatest increase in AI-related job openings? 

## Data / Visualization: 

```{r}
yearly_increase <- merged %>%
  group_by(industry, year) %>%
  summarize(yearly_total = sum(postings_count)) %>%
  arrange(industry, year)

top5 <- merged %>%
  distinct(industry, increase) %>%
  arrange(desc(increase)) %>%
  slice(1:5) %>%
  pull(industry)

yearly_increase_top5 <- yearly_increase %>%
  filter(industry %in% top5)

ggplot(yearly_increase_top5, aes(x = as.numeric(year), y = yearly_total, color = industry)) +
  geom_line(linewidth = 1.1) +
  labs(
    title = "AI Job Postings for Top 5 Growing Industries Per Year",
    x = "Year",
    y = "Number of Postings"
  )
```

```{r}
ggplot(merged, aes(x = industry, y = increase, fill = industry)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Increase in AI Job Postings by Industry",
    x = "Industry",
    y = "Increase in Job Postings"
  )
```

### To evaluate industry growth, we summarized monthly posting counts across all sectors and calculated the change between the earliest and most recent observations for each industry. We merged the trends and insights data sets based on industry to create a grouped line chart tracking year-to-year hiring trends among the top five fastest-growing industries, and a bar chart showing the total posting increase for all industries. These graphics highlight both long-term momentum and relative differences across sectors. 

### Findings: The results show that AI hiring has expanded far beyond traditional technology industries. The strongest increases occurred in Manufacturing, Education, Healthcare, Finance, and Real Estate. These sectors consistently posted more AI-related jobs over time, with strong upward trajectories in the trend lines. The pattern demonstrates that AI adoption is now embedded in industries that rely heavily on structured data, automation, predictive modeling, and process optimization. AI is becoming central to organizational strategy across the broader economy, not isolated within software firms. 

### Limitations: A limitation of the analysis is that it only shows AI related job postings, and it might not capture overall industry demand. Additionally, the method of calculating growth by comparing the first/last monthly posting counts can be sensitive to missing or irregular data, and the postings by month will lead to smoothing of seasonal fluctuations/spikes. Also, industry labels may vary, splitting related sectors and affecting comparisons. Last, although the insights dataset was cleaned, it was not fully integrated into the analysis, limiting the ability to examine factors such as salary, AI adoption, or automation risk. 

### Potential Improvements: Integrating the insights dataset would allow us to have a multivariable analysis to explore how factors like salary, AI adoption, or automation risk relate to job growth. Visualization could be enhanced by smoothing data or moving some averages to better show patterns and trends. Additionally, verifying and standardizing industry categories, detecting anomalies or outliers, and weighing postings by industry workforce size could give a better understanding of which sectors are experiencing the most significant growth in AI-related job openings. 


## Research Q3: How has AI-related hiring shifted in non-tech sectors 

## Data / Visualization: 

```{r}
insights %>%
  count(industry) %>%
  ggplot(aes(x = reorder(industry, n), y = n)) +
  geom_col(fill = "coral") +
  coord_flip() +
  labs(
    title = "Job Postings by Industry",
    x = "Industry",
    y = "Number of Postings"
  )
```

```{r}
growth_percent <- insights %>%
  filter(!is.na(job_growth_projection)) %>%
  count(job_title, job_growth_projection) %>%
  group_by(job_title) %>%
    mutate(total = sum(n)) %>%
    mutate(percentage = (n / total) * 100) %>%
  filter(job_growth_projection == "Growth", total >= 2) %>%
  arrange(desc(percentage)) %>%
  head(10)

ggplot(growth_percent, aes(x = reorder(job_title, percentage), y = percentage)) +
  geom_col(fill = "#3498db") +
  coord_flip() +
  ylim(0, 60) +
  labs(
    title = "Top 10 AI Job Roles with Highest Growth Rate",
    x = "Job Role",
    y = "% of Job Postings Labeled Growth"
  )

```

```{r}
growth_by_adoption <- insights %>%
  filter(!is.na(job_growth_projection), !is.na(ai_adoption_level)) %>%
    count(job_title, ai_adoption_level, job_growth_projection) %>%
    filter(job_growth_projection == "Growth") %>%
    arrange(ai_adoption_level, desc(n)) %>%
    group_by(ai_adoption_level) %>%
    slice_max(n, n = 5) %>%
    ungroup()
 
  ggplot(growth_by_adoption, aes(x = ai_adoption_level, y = n, color = job_title, group = job_title)) +
    geom_line() +
    geom_point(size = 2) +
    scale_x_discrete(limits = c("Low", "Medium", "High")) +
    labs(
      title = "Job Growth Trajectory by AI Adoption Level",
      x = "AI Adoption Level",
      y = "Number of Growth Postings",
      color = "Job Role"
    )
```

### We merged job postings dataset with market insights to analyze which job roles are labeled as “Growth” or “Decline”. Utilizing this information as well as “ai_adoption_level” we were able to create a chart showing posting volumes by industry, a ranking of job roles with highest percentage of “Growth” labels, and a comparison between job growth patterns across different industries and its respective AI adoption level.  

## Findings: 

### The results show that non-tech sectors show significant involvement in hiring AI-related job roles. Healthcare, Finance, and Education have the most amount of job posting. Jobs labeled “Growth” reveal the organizations outside of the tech sectors are hoping to dive deeper into the space by seeking employees with technical skills as well as people with operational or management skills. These in combination help ease the flow of AI implementation. Furthermore, the adoption level visualization demonstrates that industries with low AI adoption search to recruit for AI roles suggesting that we are at the early stage of AI usage and that investments in AI are growing. 

## Limitations: 

### The classification of roles such as “Growth” or “Decline” are not very specific measures of the market. The numbers for some sectors may also be skewed as they use more public postings while other sectors are underrepresented as they hire internally.  

## Potential Improvements: 

### In the future we could gather more datasets surrounding employee’s and companies’ exposure to AI and the skill requirements of different industries to better understand how companies in non-tech sectors utilize AI and people with AI skills differently. 

## Research Q4: What countries are leading in AI job growth relative to GDP or population? 

## Data / Visualization: 

```{r}
finalData %>%
  slice_max(jobCount, n = 15) %>%
  ggplot(aes(x = reorder(Country, jobCount), y = jobCount)) +
  geom_col(fill = "#2E86AB") +
  coord_flip() +
  labs(
    title = "Top 15 Countries by AI Job Postings (2019-2024)",
    x = NULL,
    y = "Number of Jobs"
  ) +
  theme_minimal()
```

```{r}
finalData %>%
  slice_max(jobsPerMillion, n = 15) %>%
  ggplot(aes(x = reorder(Country, jobsPerMillion), y = jobsPerMillion)) +
  geom_col(fill = "#A23B72") +
  coord_flip() +
  labs(
    title = "Top 15 Countries by AI Jobs per Million Population",
    x = NULL,
    y = "Jobs per Million People"
  ) +
  theme_minimal()
```

```{r}
finalData %>%
  filter(jobsPerBillionGdp < Inf) %>%
  slice_max(jobsPerBillionGdp, n = 15) %>%
  ggplot(aes(x = reorder(Country, jobsPerBillionGdp), y = jobsPerBillionGdp)) +
  geom_col(fill = "#F18F01") +
  coord_flip() +
  labs(
    title = "Top 15 Countries by AI Jobs per Billion USD GDP",
    x = NULL,
    y = "Jobs per Billion USD"
  ) +
  theme_minimal()
```

## Summary statistics table
```{r}
finalData %>%
  arrange(desc(jobCount)) %>%
  select(Country, jobCount, jobsPerMillion, jobsPerBillionGdp) %>%
  mutate(
    jobsPerMillion = round(jobsPerMillion, 1),
    jobsPerBillionGdp = round(jobsPerBillionGdp, 2)
  ) %>%
  head(15)
```

### We calculated AI job counts for each country and merged it with data focused on population and their economy to compare countries. We utilized 3 visuals to demonstrate the comparison. A ranking of the top countries by total job count, chart of AI jobs per million people, and another demonstrating AI jobs per billion GDP. These graphs and comparisons helped us analyze nations of varying sizes and different economic formations. 

### Findings: The U.S leads in total AI job posts, Canada, U.K, and Germany following. However, when we adjusted per capita, smaller nations like Estonia and American Samoa were higher on rankings showing that they had higher concentration of AI employment relative to the size of their workforce. When we adjusted for GDP, economies with a medium size had more generation of AI jobs per dollar of output than countries with a larger economy. This pattern showed which countries are more strategic in their investment of AI. 

### Limitations: Only 15 out of 20 countries could be matched across datasets due to inconsistent naming and missing demographic data which could have changed the result. Also, the job postings dataset from Kaggle prioritizes English speaking markets, sort of ignoring the AI powerhouses in countries like China or Japan.  

### Potential Improvements: In the future we could improve the merging by some sort of code instead of country names and add job datasets outside of just English postings. We could also use other metrics to determine why some countries have a higher AI workforce concentration than other countries. 

# 7. Discussion
----------------------------------------------------------------------------------------------------------------------------------------------------------------
### Our analysis reveals that AI's impact on employment is significant but uneven across different measures. The 47.5% salary increase from 2020-2023 shows strong demand for AI skills, while the similar hiring rates across company sizes (only 1% variation) indicate AI adoption has spread beyond just large tech companies. Industries like healthcare, finance, and real estate are leading AI job growth because they work with large amounts of data, where AI provides clear value. However, areas like automotive and transportation show flat growth, suggesting AI adoption is concentrated rather than universal. 

### The growth of both technical roles (Software Engineers, AI Researchers) and managerial positions (Operations Managers) indicates that successful AI implementation requires more than coding skills alone. Companies need people who can both build AI systems and manage their deployment. Geographic patterns show an interesting split: the United States dominates in total jobs (5,305 positions), but smaller countries like Estonia and American Samoa rank higher when adjusted for population or GDP. This suggests different countries are taking different approaches to AI development. 

### Our data has several significant limitations. First, we only capture jobs posted publicly on job boards, missing positions filled through networking or internal hiring. Second, only 15 of 20 countries could be matched between datasets due to naming differences and missing data. Third, our data heavily represents English-speaking countries and may miss significant AI activity in China, Russia, and other markets. Fourth, what counts as an AI role varies between companies, making exact comparisons difficult. Finally, our time period (2019-2024) captures recent growth but cannot show whether this represents a permanent shift or a temporary surge. 

### These limitations mean we can be confident about broad trends, such as overall salary growth and which major countries lead in AI employment, but less sure about specific numbers or rankings for smaller countries. Future research should track these patterns over longer time periods, analyze which specific AI skills are most valuable, and incorporate data from non- English job markets to get a truly global picture.

# 8. Summary 
----------------------------------------------------------------------------------------------------------------------------------------------------------------

### This analysis examined AI's impact on employment from 2019-2024 using job postings and economic data. We found four main patterns: AI salaries increased 47.5% while demand grew across all company sizes; healthcare, finance, and real estate led industry growth; both technical and managerial AI roles expanded rapidly; and geographic leadership varies by metric, with large economies dominating total jobs but smaller nations showing higher density per capita/GDP. 

### These findings matter for different groups. Workers should develop AI skills to improve career prospects, but should also focus on continuous learning as the field evolves quickly. Companies face intense competition for AI talent and must offer substantial compensation. Policymakers should note that even smaller countries can build AI capacity through focused investment, though large markets maintain scale advantages. AI represents a major workforce transformation across multiple industries. Unlike past technological changes limited to specific sectors, AI's broad applicability creates both disruption and opportunity across the economy. Our analysis captures this shift as AI moves from experimental technology to an essential business tool. The patterns we identify are not fixed - different industries and countries are adopting AI at different speeds, meaning stakeholders can influence how this transformation unfolds in their contexts. Understanding current trends provides a foundation for workers making career decisions, companies planning hiring strategies, and policymakers designing programs to support workforce adaptation.
