---
title: "AI_Analysis"
author: "Aarul"
date: "2025-11-27"
output: html_document
---

## Collaborators: Aarul Motiani, Eshaan Garg, Wayne Zhao, Finn Caughie

================================================================================
## AARUL'S RESEARCH QUESTION: How has demand for AI-related roles grown over the past 5 years?

## Loading libraries and datasets
```{r}
library(tidyverse)
library(ggplot2)
library(lubridate)

## Aarul's computer loaded datasets
job_postings <- read_delim("/Users/Aarul/UW/INFO 201/Final Project/INFO201_Final_Project/AI_Job_Posting_Trends.csv")
salary_data <- read_delim("/Users/Aarul/UW/INFO 201/Final Project/INFO201_Final_Project/data_science_salaries.csv")
```
## Data exploration for Aarul's datasets
```{r}
head(job_postings)
names(job_postings)
nrow(job_postings)
ncol(job_postings)


head(salary_data)
names(salary_data)
nrow(salary_data)
ncol(salary_data)

```

## Dataset 1 (job_postings) cleaning and organizing
```{r}
job_postings_clean <- job_postings %>%
filter(!is.na(posting_date), !is.na(job_title)) %>% 
distinct() %>% 
mutate(date = mdy(posting_date), year = year(date), month = month(date)) ## The posting date is read as a character by R, so we need to convert it into a date object
```

## Dataset 2 (market_insights) cleaning and filtering data
```{r}
salary_data_clean <- salary_data %>%
filter(work_year >= 2020, work_year <= 2023) %>%
filter(!is.na(job_title), !is.na(salary_in_usd)) %>%
distinct() %>% 
rename(year = work_year)
```

## Grouping, variable declaration, and simple statistical calculations (mean, median) for dataset 1
```{r}
job_demand <- job_postings_clean %>%
  group_by(year, job_category) %>%
```

## Grouping, variable declaration, and simple statistical calculations (mean, median) for dataset 2
```{r}
salary_trends <- salary_data_clean %>%
  group_by(year, job_category) %>%
```

## Merging both the datasets
```{r}
merged_data <- job_demand %>%
  left_join(salary_trends, by = c("year", "job_category"))
```

## Research question data analysis
```{r}

```

## Visualizations
```{r}

```

================================================================================
## FINN'S RESEARCH QUESTION: Which industries have experienced the greatest increase in AI-related job openings?
## Finn, put whatever code u have written here


#Loading and checking data
```{r}
insights <- read_delim("ai_job_market_insights.csv")
head(insights)
```


```{r}
trends <- read_delim("AI_Job_Posting_Trends.csv")
head(trends)
```

#Cleaning the data 
```{r}
insights_clean <- insights %>%
  distinct()

trends_clean <- trends %>%
  mutate(posting_date = as.Date(posting_date, format = "%m/%d/%y")) %>%
  distinct()
```

#Updating data so I can use it for merging and graphing
```{r}
trends_per_month <- trends_clean %>%
  mutate(year = format(posting_date, "%Y"), month = format(posting_date, "%m")) %>%
  group_by(industry, year, month) %>%
  summarize(postings_count = n()) #counts number of rows per group
```

```{r}
industry_increase <- trends_per_month %>%
  group_by(industry) %>%
  summarize(first = first(postings_count), last = last(postings_count), increase = last - first) %>%
  arrange(desc(increase))
```

```{r}
head(trends_per_month)
head(industry_increase)
```

#Merging the data
```{r}
merged <- merge(industry_increase, trends_per_month, by = "industry")
head(merged)
```

```{r}
merged <- merged %>%
  arrange(desc(increase))

```

#This graph shows the job postings for the top 5 growing industries over the years
```{r}
yearly_increase <- merged %>%
  group_by(industry, year) %>%
  summarize(yearly_total = sum(postings_count)) %>%
  arrange(industry, year)

top5 <- merged %>%
  distinct(industry, increase) %>%
  arrange(desc(increase)) %>%
  slice(1:5) %>%
  pull(industry)

yearly_increase_top5 <- yearly_increase %>%
  filter(industry %in% top5)

ggplot(yearly_increase_top5, aes(x = as.numeric(year), y = yearly_total, color = industry)) +
  geom_line(linewidth = 1.1) +
  labs(
    title = "AI Job Postings for Top 5 Growing Industries Per Year",
    x = "Year",
    y = "Number of Postings"
  )
```


#This graph uses merged data to show increase of AI job postings by industry per year 
```{r}
ggplot(merged, aes(x = industry, y = increase, fill = industry)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Increase in AI Job Postings by Industry",
    x = "Industry",
    y = "Increase in Job Postings"
  )
```


================================================================================
## WAYNE'S RESEARCH QUESTION: How has AI-related hiring shifted in non-tech sectors?
## Wayne, decide what code u need here and remove what you're not using

```{r}
market_insights <- read_delim("./ai_job_market_insights.csv")
```

```{r}
lowercase_clean <- function(df) {
  names(df) <- tolower(names(df))
  df
}

job_postings <- job_postings %>%
  rename(location = company_location)

market_insights <- lowercase_clean(market_insights)

names(job_postings)
names(market_insights)

merged_data <- job_postings %>%
  left_join(market_insights, by = "job_title", relationship = "many-to-many")

merged_data <- merged_data %>%
  select(-ends_with(".y"))

merged_data <- merged_data %>%
  mutate(
    remote_friendly = ifelse(remote_category == "Fully Remote", "Yes", remote_friendly),
    remote_friendly = ifelse(remote_category == "On-site", "No", remote_friendly),
    remote_friendly = ifelse(remote_category == "Hybrid", "Hybrid", remote_friendly)
      )

merged_data
```

## columns needed for Q3
```{r}
job_postings %>%
  select(industry, posting_date, job_title, salary_usd, company_size, company_location, required_skills)

market_insights %>%
  select(industry, ai_adoption_level, automation_risk, job_growth_projection, salary_usd, required_skills, job_title)
```

```{r}
market_insights <- market_insights %>%
  mutate(
    company_size = ifelse(company_size == "Small", "S", company_size),
    company_size = ifelse(company_size == "Medium", "M", company_size),
    company_size = ifelse(company_size == "Large", "L", company_size)
    ) %>%
  mutate()

market_insights
job_postings

```


================================================================================
## ESHAAN'S RESEARCH QUESTION: What countries are leading in AI job growth relative to GDP or population?







